Class-based network segregation, Economic Inequality and Redistributive Preferences across societies

### 2024-07-19 paper1_jiturra.Rmd
up humboldt

 During the past decades, cross-national studies on redistributive preferences and social class have predominantly focused on individuals and households [@lindh_class_2020]. In contrast, less attention has been paid to the role of social networks on redistributive preferences. Sociologists have demonstrated that individuals tend to form segregated social networks according to social class [@mcpherson_birds_2001;@otero_lives_2022]. Additionally, attitudes toward economic inequality can be affected by class segregation, as it implies distance from other social classes'  economic conditions, lifestyles, and worldviews [@lindh_missing_2021;@lee_consider_2023;@paskov_crossclass_2022]. Therefore, focusing on social ties can improve the understanding of the relationship between social class and redistributive preferences, as it broadens the scrutiny of class relations in terms of internalization of shared norms, class identity formation, and collective solidarity. [@Svallfors2006;@goldthorpe_service_1982]. In this discussion, the first part of this study examines the micro-level relationship between class-based network segregation and redistributive preferences. Subsequently, I focus on how cross-national differences in income inequality can moderate the relation of network segregation and support for redistribution.

@@

As network segregation can widen the class divide in redistributive preferences, little is known about how income inequality may affect the association between class-based network segregation and support for redistribution. On the one hand, studies on the class-attitude link suggest that income inequality is crucial for understanding how class relations are reflected in redistributive demands, as it represents the current state of distributive class struggle in contemporary capitalist societies [@edlund_democratic_2015;@curtis_how_2015]. A consistent finding in this literature is that the upper classes tend to be more supportive of redistribution in contexts of high inequality, while the already high redistributive demands of the working class remain stable regardless of inequality levels [@sachweh_why_2019;@dimick_altruistic_2017]. On the other hand, income inequality has implications for social participation and network composition. Indeed, inequality reinforces stratified access to social activities and widens the distance between classes, resulting in a higher diversity of networks and activities in the upper classes and more marginalization of the lower classes [@lancee_income_2012;@otero_differences_2023;@pichler_patterns_2007]. However, current efforts to examine the impact of income inequality have focused primarily on social networks or support for redistribution. Thus, a key question that remains unanswered is the extent to which higher income inequality moderates the relationship between social class-based network segregation and redistributive preferences.

@@

Over the past decades, the study of political attitudes in industrialized societies has consistently demonstrated the relevance of social class as a driver of public opinion [@lindh_class_2020, p. 421]. In this view, social class not only represents individuals' labor market relations but also their economic interests and moral perspectives on the role of the market and the state in the distribution and redistribution of resources [@Svallfors2006]. Here, redistributive preferences are understood as the support for policies and mechanisms aimed at reducing economic inequality through the redistribution of wealth and resources [@mccall_americans_2009]. These preferences encompass views on taxation, welfare programs, public services, and other government interventions designed to transfer resources from wealthier to less wealthy individuals or groups in society [@garcia-sanchez_two_2022].

Class-based explanations of redistributive preferences have primarily focused on the individual or household level but are not exclusively limited to these contexts. It has been argued that redistributive preferences can be explained through labor market situations, which encompass access to economic resources and risk exposure [@meltzer_rational_1981;@rehm_risks_2009]. Additionally, while material interests may dominate in situations of scarcity, value-driven motivations become a significant explanatory factor for redistributive preferences under conditions of greater certainty and are weaker under material precariousness [@Kulin2013;@Maldonadoetal2019]. Other approaches suggest that the substantial time workers spend on their jobs allows social relations in the workplace to imprint normative views that ultimately shape political opinions [@oesch_redrawing_2006]. For example, the continuous and diverse social interactions inherent in interpersonal services can foster empathy and reinforce egalitarian values [@kitschelt_occupations_2014]. In contrast, vertical monitoring in managerial occupations and the emphasis on autonomy in self-employed roles tend to strengthen self-interested and conservative political views [@oesch_electoral_2018].

Empirically, the current understanding of class divides in redistributive preferences is extensive [@lindh_public_2015;@curtis_how_2015;@langsaether_more_2020;@brooks_why_2010]. Nevertheless, considering class-based network ties highlights the relevance of the class situation, as it encompasses economic interests and life chances that drive communal action beyond individuals [@weber_class_2011, pp. 57-59]. Hence, considering the structure of social ties allows one to look more comprehensively into how class relations structure redistributive preferences.

@@

Class relations can be understood as the structure of social ties between different classes within the social structure, represented by networks formed by various social classes [@blau_macrosociological_1977]. Empirical evidence consistently shows homophily in social relations within the social network literature [@mcpherson_birds_2001]. This indicates that friendships and family ties are often homogeneous regarding social status and demographic characteristics, whereas distant ties connect individuals to different social groups and enhance network diversity [@diprete_segregation_2011;@lazarsfeld_friendship_1954]. Moreover, socialization preferences do play a role in the formation of segregated networks [@visser_attitudes_2004; @homans_human_1951]. However, the similarity of attitudes within these networks arises more from the social differentiation structuring class ties than from individual preferences [@feld_focused_1981].

@@

How does social class stratify social networks? Empirical studies on network diversity reveal that higher civic engagement in formal organizations increases the likelihood of the upper class forming connections with diverse individuals, contrasting with the more homogeneous participation observed among the working class [@pichler_social_2009]. Similar patterns are found in the composition of social ties, where the upper and intermediate classes maintain more diverse and prestigious social environments compared to the working class [@cepic_how_2020; @carrascosa_class_2023]. In contrast, studies on network segregation suggest that property-based boundaries are less permeable than authority-based ones. For example, @wright_relative_1992 suggest that class interests widen the social distance between proprietors and manual workers, while the intermediate position of supervisors and their higher contact frequency with manual workers make the formation of friendship ties more likely. Similarly, @otero_open_2021 found that the intermediate class exhibits greater permeability compared to the more homogeneous networks of the working class. This suggests that limited life chances and lower capacity for social engagement ultimately result in segregation. Conversely, the upper class is more homogeneous because it tends to self-select, a practice that ultimately seeks to reproduce its privileged positions [@otero_open_2021].

>>>>

During the past decades, cross-national studies on social class and redistributive preferences have predominantly focused on individuals and households [@lindh_class_2020]. In contrast, less attention has been paid to how the structure of social relations can contribute to explaining class differences in redistributive preferences. In this regard, sociologists have pointed out that a structural feature of contemporary societies is the tendency of social classes to form segregated social networks [@mcpherson_birds_2001;@otero_lives_2022]. Furthermore, recent studies have argued that class segregation can indeed affect redistributive preferences, as it implies lower social integration and detachment from the economic conditions, lifestyles and worldviews of other social classes [@lindh_missing_2021;@lee_consider_2023;@paskov_crossclass_2022]. Consequently, focusing on the structure of social networks can contribute to a better understanding of the link between class relations and redistributive preferences. Concerning this, networks broaden the scope of class analysis through class-specific relational aspects, such as shared norms internalization, identity formation, and collective solidarity [@Svallfors2006;@goldthorpe_service_1982]. Adding to this debate, this research starts by examining the micro-level relationship between segregated class-based networks and redistributive preferences. Then, I scrutinize how cross-national differences in income inequality may moderate the relationship between network segregation and redistributive preferences.

@@

Since the degree of segregation may amplify already stratified class differences in redistributive preferences, little is known about how income inequality may moderate this relationship. On the one hand, studies on the class-attitude link suggest that income inequality is crucial for understanding how class relations are reflected in redistributive demands, as it represents the current state of distributive class struggle in contemporary capitalist societies [@edlund_democratic_2015;@curtis_how_2015]. A consistent finding in this literature is that the upper classes tend to be more supportive of redistribution in contexts of high inequality, while the already high redistributive demands of the working class remain relatively stable regardless of inequality levels [@sachweh_why_2019;@dimick_altruistic_2017]. On the other hand, it is well documented that income inequality has consequences on social relations. Indeed, it can reinforce stratified access to social activities and widen the distance between classes, which means that the upper classes participate more and hold more diverse networks, while the lower classes tend to be more inactive and segregated as inequality rises [@lancee_income_2012;@otero_differences_2023;@pichler_patterns_2007]. However, current efforts to examine the impact of income inequality have focused on either social networks or support for redistribution. Thus, a key question that remains unanswered is the extent to which income inequality moderates the relationship between social class-based network segregation and redistributive preferences.

@@

Over the past decades, the study of political attitudes in industrialized societies has consistently demonstrated the relevance of social class as a driver of public opinion [@lindh_class_2020]. In this view, social class not only represents individuals' labor market relations but also their economic interests and moral perspectives on the role of the market and the state in the distribution and redistribution of resources [@Svallfors2006]. Here, redistributive preferences are understood as the support for policies and mechanisms aimed at reducing economic inequality through the redistribution of wealth and resources [@mccall_americans_2009]. These preferences encompass views on taxation, welfare programs, public services, and other government interventions designed to transfer resources from wealthier to less wealthy individuals or groups in society [@garcia-sanchez_two_2022].

Class-based explanations of redistributive preferences have primarily focused on the individual or household level but are not exclusively limited to these contexts. It has been argued that redistributive preferences can be explained through labor market situations, which encompass access to economic resources and risk exposure [@meltzer_rational_1981;@rehm_risks_2009]. Additionally, while material interests may dominate in situations of scarcity, value-driven motivations become a significant explanatory factor for redistributive preferences under conditions of greater certainty and are weaker under material precariousness [@Kulin2013;@Maldonadoetal2019].

Other approaches have emphasized that social relations in the workplace can imprint normative views that ultimately shape political opinions due to the substantial time workers spend on their jobs [@oesch_redrawing_2006]. For example, the continuous and diverse social interactions inherent in interpersonal services can foster empathy and reinforce egalitarian values [@kitschelt_occupations_2014]. In contrast, vertical monitoring in managerial occupations and the emphasis on autonomy in self-employed roles tend to strengthen self-interested and conservative political views [@oesch_electoral_2018].

Empirically, the current understanding of class divides in redistributive preferences is extensive [@lindh_public_2015;@curtis_how_2015;@langsaether_more_2020;@brooks_why_2010]. Nevertheless, I suggest that focusing on network ties stresses the relevance of the _class situation_ of individuals and social surroundings, as it encompasses economic interests and life chances that drive communal action  [@weber_class_2011, pp. 57-59]. Hence, studying connections between classes from a network perspective allows one to examine how class relations structure redistributive preferences comprehensively.

@@

Class relations can be understood as the structure of social ties between different classes within the social structure, represented by networks formed by various social classes [@blau_macrosociological_1977]. Empirical evidence in the social network literature consistently demonstrates that homophily is a socially structured tendency in social relations [@mcpherson_birds_2001]. For instance, this implies that friendships and family ties are often homogeneous concerning social status or demographic characteristics, whereas distant ties link individuals to different social groups and raise network diversity [@diprete_segregation_2011;@lazarsfeld_friendship_1954]. At the same time, it is well established that socialization preferences play a role in forming segregated networks [@visser_attitudes_2004; @homans_human_1951]. However, it has been argued that attitude similarity within these networks arises more due to the contact opportunities that structure the class composition of networks than individual sociability preferences [@feld_focused_1981].

@@

How does social class stratify social networks? Empirical studies on network diversity reveal that higher civic engagement in formal organizations increases the likelihood of the upper class forming connections with diverse individuals, contrasting with the more homogeneous participation observed among the working class [@pichler_social_2009]. Similar patterns are found in the composition of social ties, where the upper and intermediate classes maintain more diverse and prestigious social environments compared to the working class [@cepic_how_2020; @carrascosa_class_2023]. In contrast, studies on network segregation suggest that property-based boundaries are less permeable than authority-based ones. For example, @wright_relative_1992 suggest that class interests widen the social distance between proprietors and manual workers, while the intermediate position of supervisors and their higher contact frequency with manual workers make the formation of friendship ties more likely. Similarly, @otero_open_2021 found that the intermediate class exhibits greater permeability compared to the more homogeneous networks of the working class. This suggests that limited life chances and a lower capacity for social engagement ultimately result in segregation. Conversely, the upper class is more homogeneous because it tends to self-select, a practice that ultimately seeks to reproduce its privileged positions [@otero_open_2021].
